---
title: "Exponential Distribution Simulation"
author: "MattDSquared"
date: "Saturday, June 20, 2015"
output:
  word_document: default
  html_document:
    keep_md: yes
---
As part of the Statistical Inference class on Coursera, this project investigates the exponential distribution in R and compares it with the Central Limit Theorem. The mean of an exponential distribution is 1/lambda and the standard deviation is also 1/lambda.

## Project Setup
```{r setup, results='hide'}
library(ggplot2)
setwd("~/../datascience/StatisticalInferenceProject")

lambda <- .2
n <- 1000
n_samples <- 40
```

## Distribution of `r n` expontially distributed samples
Using a simulation of `r n` samples, the original exponential distribution for paramaeter lambda set to `r lambda` looks like this:

```{r distr.simulate}
set.seed(35)
samples <- rexp(n, lambda)
```
```{r distr.display, fig.height=3.5, fig.width=6}
meanlabel <- paste(" mean=",round(mean(samples),2),sep="")
gg <- qplot(samples) + 
    geom_vline(xintercept=mean(samples), color="red") + 
    geom_text(aes(x=mean(samples), y=0, label=meanlabel),
              colour="red", vjust=1, hjust=0, size=3) +
    labs(x="sample values") + 
    labs(title=paste("Histogram of", n, "samples from the Exponential",
                     "Distribution"))
print(gg)
```

Note the mean does in fact approximate the theoretical mean of 1/lambda = `r 1/lambda`.

## Distribution of `r n` sample means
We can now compare the exponential distribution with the distribution of its sample means. The following simulation builds an exponential distribution from `r n` sample means, with sample size of `r n_samples`. 
```{r means.simulate}
means <- NULL # bad memory management, consider preallocating
set.seed(19)
for (i in 1:n) means <- c(means, mean(rexp(n_samples,lambda)))
```

```{r means.display, fig.height=3.5, fig.width=6}
mean_means <- mean(means)
sd_means <- mean(means) + c(-1,1)*sd(means)

meanlabel <- paste(" mean=", round(mean_means,2),sep="")
sdlabel <- c(paste(" sd=",round(mean_means,2), "-", 
                   round(sd(means),2),sep=""), 
             paste(" sd=",round(mean_means,2), "+", 
                   round(sd(means),2),sep=""))
df <- data.frame(x = means)
gg <- ggplot(df, aes(x = x)) + 
    geom_histogram(aes(y = ..density..)) +
    geom_vline(xintercept=mean_means, color="red") + 
    geom_text(aes(x=mean(means), y=0, label=meanlabel),
              colour="red", vjust=1, hjust=0, size=3) +
    geom_vline(xintercept=sd_means, color="blue") + 
    geom_text(aes(x=sd_means[1], y=0, label=sdlabel[1]),
              colour="blue", vjust=1, hjust=0, size=3) + 
    geom_text(aes(x=sd_means[2], y=0, label=sdlabel[2]),
              colour="blue", vjust=1, hjust=0, size=3) + 
    stat_function(geom="line", fun=dnorm, colour = "green",
                  arg=list(mean=mean(means), sd=sd(means))) +
    labs(x="sample values") + 
    labs(title=paste("Histogram of", n, "sample means \nfrom the Exponential",
                     "Distribution with sample size of", n_samples))
print(gg)
```

Exponential distributions do in fact follow the central limit theorem. Compared to the previous plot, the above plot shows an approximately gaussian distribution (shown in green). 

The sample mean is `r round(mean(means),2)` which is a very close approximation to the theoretical mean of the distribution:
```{r}
1/lambda
```
The sample standard deviation is `r round(sd(means),3)` which is also a very close approximation of the theoretical standard deviation of the distribution:
```{r}
round((1/lambda)/sqrt(n_samples),3)
```
